Verifying Spatial Properties of Stencil Computations

### 2016-11-16 paper.tex
FREQUENCY TABLE

\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
$\forall S,
  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus, our model validates specification synthesis.
%
\subsection{Step 5: Simplification}
The final step applies a simplification pass, orienting the
equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
based on some simple heuristics.
For our example, we infer and then synthesise the
final 5-point specification:
%%
\begin{minted}[linenos=false,breakindent=2em]{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
\end{minted}
%%

@@

We would not expect to infer a stencil for each array statement we
found because our analysis restricts 
the array-subscript-statements that we class as a
stencil. In fact, we was able to infer a stencil from 30\% of
these statements. A single statement can involve multiple arrays and
we ended up with 60,525 specifications. This shows that we can express
a large number of stencil uses within our high-level abstraction and
validates our initial hypothesis that many stencil computations have a
regular shape.

There were a variety of reasons why we did not infer a
stencil from every potential statement: \\
\noindent 
1) \textbf{Non-subset induction variables} occur when the
induction variables on the RHS are not a subset of those in the LHS. These
cases are not stencils by our definition. The degenerate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators \eg{} computing the sum over an array; \\
\noindent
2) \textbf{Derived induction variables} where an
index (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};  \\
\noindent
3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

@@

\textbf{Bounded} specifications were fairly unusual, with only 157
\texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
always also paired with an upper bound.

@@

The single- and multi-action classes represent more complex stencils
with a real possibility for programmer error. For example, the Unified
Model has an implementation of the Smagorinsky subgrid-scale model for
calculating turbulence on which our inference yields 39 specs from 340
lines of code. This is an impressive reduction given
the complexity of the algorithm.  We show one example in~\Cref{fig:smagorinsky}. This specifies the access pattern to
a 3-dimensional array (which we have renamed to be called
\texttt{x}) arising from a kernel of 93 lines of code involving 142
individual array subscripts. The specification was the most complex
seen in our corpus, yet it still represents a significant abstraction
to the spatial behaviour given size and complexity of the kernel it describes.

\subsection{Detecting errors in the 2-D Jacobi iteration}
\label{sec:jacobi}
One common example of a stencil computation is two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
is given by:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and we infer a precise specification for this, as:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
\end{minted}
%
To test our implementation, 
we examined whether programmer errors would be detected by replacing
the array index offsets with $-1$, $0$, or $1$ and running our
verification algorithm. Our checking procedure correctly reported a verification
failure in each of 6,537 permutations corresponding to an error.  
The iteration computes the average of four adjacent
cells so $24$ ($4$ factorial) of the possible array index perturbations
are correct, all of which are accepted by checking.

@@

Various deductive verification tools can express array indexing
in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
can be given for a stencil computation but replicating the fine-grained
indexing seen in the code and therefore similarly prone to indexing
errors. Our approach is much more abstract-- it does
not aim to reify indexing in the specification, but
provides simple spatial descriptions which capture a large number
of common patterns.

@@

% GPGPU
Our work has similarities with efforts to verify kernels
written for General-Purpose GPU
programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
%kernel, and GPGPU programming can be viewed as a massively parallel
%method of transforming a large matrix. 
However, their focus is
mainly on the synchronisation of kernels and the avoidance of
data races, while we are interested in correctness 
embedded within a more typical general-purpose programming
language. %Other work on GPGPU computation, such as
%\citet{Zhang:2012:CGO}, has focused primarily on generating
%optimised code based on relatively simple specifications: to
%provide performance while keeping the programmer's effort within
%reason.
%
% Sketching Stencils
\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
unoptimised ``reference'' stencils, coupled with partial
implementations which are completed by a code
generation tool. %These kinds of specifications are just simple
%implementations, so this tool is useful for hand-written, optimised
%stencils. 
The primary purpose of this tool is optimisation rather than
correctness, and the language of specification is much different and
more elaborate than in our approach.

% Pochoir
\citet{Tang:2011:SPAA} define a specification language for writing stencils
embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that are then
compiled into parallel programs based on trapezoidal decompositions with
hyperspace cuts. The Pochoir specifications are used for describing the kernel,
boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
reluctant to implement the high-performance cache-oblivious
``hypertrapezoidal'' algorithms for stencils.  Like much of the related work,
the goal is optimisation rather than verifying program correctness.

By contrast, the work of \citet{Abe:2013:IPDPSW} studies
correctness explicitly by bringing a form of model-checking to
verify certain stencil computations in the context of
parallel implementations 
 in partitioned global address
space languages. %These are scenarios where an array is divided into
%subarrays on multiple processors but each has the ability to access
%the others' memory. The authors avoid the state explosion dilemma of
%model-checking by relying upon the fact that accesses to the
%``boundary elements'' will require a different method by virtue of the
%non-locality of that memory. Since this access is performed in a
%different manner, those can be identified automatically by static
%analysis. To achieve this, 
Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
effort goes towards describing the distribution of the
computation over multiple processors. The code for the stencil kernel
is generated from a relatively high-level specification. 
In contrast, we integrate directly
into existing, legacy codebases and established languages, bringing
the benefits of verification more easily to scientific computing.%such as
%older versions of Fortran.% We infer stencil specifications from
%Fortran code and check annotations on stencil computations within
%Fortran code.

>>>>

\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
$\forall S,
  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus, our model validates specification synthesis.

The final step applies a simplification pass, orienting the
equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
based on some simple heuristics.
For our example, we infer and then synthesise the
final 5-point specification:
%%
\begin{minted}[linenos=false,breakindent=2.4em,xleftmargin=0.2em]{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
\end{minted}
%%

@@

We would not expect to infer a stencil for each array statement we
found because our analysis restricts 
the array-subscript-statements that we classify as a
stencil. In fact, we were able to infer a stencil from 30\% of
these statements. A single statement can involve multiple arrays and
we ended up with 60,525 specifications. This shows that we can express
a large number of stencil shapes within our high-level abstraction and
validates our initial hypothesis that many stencil computations have a
regular shape.

@@

\textbf{Bounded} specifications occured with 157
\texttt{atMost} bounds and 36 \texttt{atLeast}, the latter of which were
always also paired with an upper bound.

@@

The single- and multi-action classes represent more complex stencils
with a real possibility for programmer error. As an extreme example, the Unified
Model has an implementation of the Smagorinsky subgrid-scale model for
calculating turbulence on which our inference yields 39 specs from 340
lines of code. This is an large reduction given
the complexity of the algorithm.  We show one example
in~\Cref{fig:smagorinsky} which specifies the access pattern to
a 3-dimensional array (which we have renamed to 
\texttt{x}) arising from a kernel of 93 lines of code involving 142
array subscripts. The specification was the most complex
seen in our corpus, yet it still represents a significant abstraction
of the spatial behaviour given size and complexity of the kernel it describes.

We measured the frequency at which individual specifications involved 
multiple occurences of \term{+} and \term{*}:

\vspace{0.25em}
\setlength{\tabcolsep}{0.57em}
{\small{
\hspace{-1em}\begin{tabular}{c|cccccccc}
& 0 & 1 & 2 & 3 & 4 & 5 & 6 & $> 7$ \\ \hline
\term{+} & 60,265 & 220 & 15 & 12 & 5 & 4 & 4 & 0 \\
\term{*} & 31,393 & 15,151 & 13,457 & 424 & 41 & 42 & 5 & 12
\end{tabular}}}
\vspace{0.2em}

\noindent
\eg{}, roughly half the specifications did not involve $\term{*}$, 
a quarter use one $\term{*}$ and just under a quarter
use two $\term{*}$ operators.

\paragraph{Limitations}

There were various reasons why we did not infer 
specifications on every looped array computation: \\
\noindent 
1) \textbf{Non-subset induction variables} occur when the
induction variables on the RHS are not a subset of those in the LHS. These
cases are not stencils by our definition. The degenerate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators \eg{} computing the sum over an array; \\
\noindent
2) \textbf{Derived induction variables} where an
index (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};  \\
\noindent
3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

\subsection{Detecting errors in the 2-D Jacobi iteration}
\label{sec:jacobi}
One common example of a stencil computation is the two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
is given by:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
We infer a precise specification of its shape as:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
\end{minted}
%
To test our implementation, 
we examined whether programmer errors would be detected by replacing
the array index offsets with $-1$, $0$, or $1$ and running our
verification algorithm. Our checking procedure correctly reported a verification
failure in each of 6,537 permutations corresponding to an error.  
The iteration computes the average of four adjacent
cells so $24$ ($4$ factorial) of the possible array index perturbations
are correct, all of which are accepted by our checker.

@@

Various deductive verification tools can express array indexing
in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
can be given for a stencil computation but must use fine-grained
indexing as in code and therefore is similarly prone to indexing
errors. Our approach is much more abstract-- it does
not aim to reify indexing in the specification, but
provides simple spatial descriptions which capture a large number
of common patterns.

@@

% GPGPU
Our work has similarities with efforts to verify kernels
written for General-Purpose GPU
programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
%kernel, and GPGPU programming can be viewed as a massively parallel
%method of transforming a large matrix. 
However, their focus is
mainly on the synchronisation of kernels and the avoidance of
data races., while we are interested in correctness 
%embedded within a more typical general-purpose programming
%language. %Other work on GPGPU computation, such as
%\citet{Zhang:2012:CGO}, has focused primarily on generating
%optimised code based on relatively simple specifications: to
%provide performance while keeping the programmer's effort within
%reason.
%
% Sketching Stencils
\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
unoptimised ``reference'' stencils, coupled with partial
implementations which are completed by a code
generation tool. %These kinds of specifications are just simple
%implementations, so this tool is useful for hand-written, optimised
%stencils. 
The primary purpose of this tool is optimisation rather than
correctness, and the language of specification is 
more elaborate than ours.

% Pochoir
\citet{Tang:2011:SPAA} define a specification language for writing stencils
embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that are then
compiled into parallel programs based on trapezoidal decompositions with
hyperspace cuts. Pochoir specifications are used for describing the kernel,
boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
reluctant to implement the high-performance cache-oblivious
``hypertrapezoidal'' algorithms.  Like much of the related work,
the goal is optimisation rather than verifying program correctness.

By contrast, the work of \citet{Abe:2013:IPDPSW} studies
correctness bringing a form of model-checking to
verify certain stencil computations in the context of
parallelism 
 in partitioned global address
space languages. %These are scenarios where an array is divided into
%subarrays on multiple processors but each has the ability to access
%the others' memory. The authors avoid the state explosion dilemma of
%model-checking by relying upon the fact that accesses to the
%``boundary elements'' will require a different method by virtue of the
%non-locality of that memory. Since this access is performed in a
%different manner, those can be identified automatically by static
%analysis. To achieve this, 
Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
effort goes towards describing the distribution of the
computation over multiple processors. The code for the stencil kernel
is generated from a relatively high-level specification. 
In contrast, we integrate directly
into existing, legacy codebases and established languages, bringing
the benefits of verification more easily to scientific computing.%such as
%older versions of Fortran.% We infer stencil specifications from
%Fortran code and check annotations on stencil computations within
%Fortran code.
